### 2019-04-10 Manuscript/Manuscript.tex
Updates from Overleaf

One of the hypothesized effects of the emplacement of large igneous provinces (LIPs) is planetary cooling on million-year timescales associated with enhanced silicate weathering. This study combines reconstructions of the original surface extent of LIPs through the Phanerozoic, a paleogeographic model, and a paramaterization of LIP erosion to estimate LIP area in all latitudinal bands through the Phanerozoic. This analysis reveals no significant correlation between total LIP area, nor LIP area in the tropics, and the extent of continental ice sheets. The largest peaks in tropical LIP area are at times of non-glacial climate. These results suggest that changes in planetary weatherability associated with LIPs are not the fundamental control on whether Earth is in a glacial or non-glacial climate, although they could provide a secondary modulating effect in conjunction with other processes.

@@

Global weatherability is the sum of factors aside from climate itself that contribute to overall global weathering and associated CO$_2$ consumption, such as the latitudinal distribution of continents and mountain belts. On a planet with high weatherability, the CO$_2$ input from volcanism can be removed via silicate weathering at a lower atmospheric CO$_2$ concentration than on a less weatherable planet. Basaltic regions consume more CO$_2$ than regions where bedrock composition is closer to bulk continental crust because of the relatively high concentrations of Ca and Mg (that ultimately sequester carbon through precipitation as carbonate) and the high weathering rates of mafic lithologies \citep{Dessert2003a}. Data from basaltic watersheds show that chemical weathering rates are greatest in regions with high runoff. As a result, CO$_2$ consumption in basaltic regions is most pronounced in the tropical rain belt \citep{Dessert2003a, Hartmann2009a, Hartmann2014a}.

One aspect of large igneous province (LIP) emplacement that has been hypothesized to relate to long-term global climate is the effect such emplacement could have on increasing global weatherability and driving cooling. In particular, the emplacement of LIPs in the tropics has been hypothesized to be associated with specific episodes of climatic cooling on Earth. In the Neoproterozoic Era, the emplacement of the ca. 720~Ma Franklin LIP in the tropics, in concert with elevated runoff rates associated with continental break-up, has been implicated as a major contributor to the cooling that initiated the Sturtian `Snowball Earth' \citep{Donnadieu2004b, Macdonald2010a, Cox2016a}. In the Cenozoic, the movement of the Deccan LIP into the tropical rain belt, in concert with the low-latitude emplacement of the Ethiopian Traps, has been implicated in drawing down CO$_2$ levels in the lead-up to Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}. More recently, \citet{Johansson2018a} used a paleogeographic model to suggest that tropical LIP area correlates with Phanerozoic climate change through comparisons to \textit{p}CO$_{2}$ proxy records.

@@

This study combines reconstructions of the original surface extent of LIPs through the Phanerozoic, a paleogeographic model, and a parameterization of LIP erosion to estimate LIP area in all latitudinal bands through the Phanerozoic. We then compare these LIP area time series to the latitudinal extent of continental ice sheets - a proxy for Earth's long term climate state.

Outlines of the original surface extent of continental LIPs through the Phanerozoic (Fig. \ref{fig:LIP_map}) were slightly modified (to ensure that all currently exposed volcanic lithologies are encapsulated by the initial LIP area polygons) from the compilation of \citet{Ernst2017a} and \citet{Ernst2019a}, and emplacement ages were taken from the literature (Table \ref{tab:LIPs}). The LIP original surface extent compilation seeks to reconstruct the original surface extent of LIPs with the caveat that there can be significant uncertainty with doing so, particularly for older more deeply eroded LIPs. These polygons encapsulate all of the rocks associated with a given LIP, including dikes, sills, and layered intrusions, in addition to subaerial volcanics (Fig. \ref{fig:LIP_map}). For some LIPs, this approach may lead to an overestimate of original surface extent, given that subsurface intrusions could extend over a broader area than the surface volcanics. The polygons also assume complete surface coverage between wide-spread remnants, creating further potential for these outlines being over-estimates. However, despite these uncertainties, this approach likely provides the best estimates available of original extent for ancient LIPs. The extents of presently-exposed volcanics associated with LIPs that were used for present-day area estimates, and the construction of the original extent polygons by \citet{Ernst2017a} and \citet{Ernst2019a}, were taken from a number of resources including the PLATES compilation \citep{Coffin2006a} and more recent compilation efforts associated with the LIPs Reconstruction Project (\citealp{Ernst2013a}; Table \ref{tab:LIPs}).

After LIPs are emplaced, they will progressively erode. In order to account for the associated decrease in area with time, \citet{Godderis2017a} took the approach of fitting an exponential decay function to estimates of the original surface extent and the current surface extent for 5 LIPs. They used the resulting exponential decay constants to develop a first-order parameterization to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which there are estimates of the original surface extent of the province and the current surface extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate first-order representation of the progressive reduction in LIP area (Fig. \ref{fig:LIP_preservation}). The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our models (see below), we prefer to exclude from our estimation of a representative LIP area half-life the 6 of 19 LIPs which have been partially or completely buried. This latter approach yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). Although the exponential fit to the 13 unburied LIPs is good (i.e. it yields a low root mean square error of 0.14), if each LIP is fit individually with an exponential decay function, there is variability in the estimated half-lives such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps (Table \ref{tab:LIPs}). In the analysis of LIP area through time, we implement decay scenarios informed by this analysis. Scenario `t$_{1/2}$ = 36~Myr' uses the best-fit half-life of 36~Myr, while `t$_{1/2}$ = 120~Myr' uses the slower decay. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. The LIP reconstructions used in this study also include pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs does not significantly influence the calculated LIP areas throughout the Phanerozoic, which is the focus of this analysis.

Of the tectonic factors that could alter exposed LIP area, the most consequential is near-immediate burial by sediments of LIP volcanics that are co-located with a rift basin. There are numerous examples in the record where there is partial or complete burial of a LIP associated with rifting and thermal subsidence (Table \ref{tab:LIPs}). For example, the Afar LIP is both associated with the Ethiopian Traps which form plateau flood basalts as well as successful rifting in the region of the Red Sea that has resulted in burial (Fig. \ref{fig:LIP_map}). To account for the rapid decrease in exposed surface area that would result from burial by sediments in a rift basin, we impose two different burial scenarios for LIPs that are co-located with rifting. The `50\% burial' scenario imposes instantaneous burial of 50$\%$ of the LIP area while the `100\% burial' imposes instantaneous burial of the entire LIP as an end-member scenario. A limitation of our treatment of LIPs that are co-located with rifting is that we instantaneously `bury' all of these LIPs to the same degree, when in fact the degree/timing of burial of these LIPs may vary substantially. The LIP area analysis uses all of the available combinations of the distinct decay and burial scenarios described above.

@@

The original extent LIP polygons were assigned a plate ID corresponding to a tectonic unit on Earth using the polygons of \citet{Torsvik2016a} for the Phanerozoic. The LIP polygons and tectonic units were reconstructed from 520~Ma to the present utilizing the paleogeographic model of \citet{Torsvik2016a} in the spin axis reference frame (anchor plate ID of 1). This paleogeographic model was updated to include revisions to Ordovician Laurentia \citep{Swanson-Hysell2017a} and Paleozoic Asia \citep{Domeier2018a}. Reconstructions and area calculations within latitude bands utilized the pyGPlates function library and custom Python scripts. The total LIP area and the LIP area reconstructed within the tropical rain belt were calculated for the various decay and burial scenarios at a resolution of 5~Myr (Fig. \ref{fig:LIP_Areas}B and C). All the data and code necessary to reproduce the analyses and figures presented in this study can be downloaded from GitHub: \url{https://github.com/Swanson-Hysell-Group/Large_Igneous_Provinces}.

We use 15\textdegree~S to 15\textdegree~N as the definition of the tropical rain belt, as those latitudes approximately correspond with a sharp increase in zonal mean precipitation to values greater than 1.0~m/yr in modern climatalogical data \citep{Kalnay1996a}. In \citet{Evans2006a}, the reconstructed paleolatitudes of basins with thick, basin-wide evaporite deposition are shown to be consistently in the subtropics throughout the Phanerozoic (and potentially beyond), suggesting the stability of the large-scale atmospheric circulation that gives rise to intense precipitation in the tropical rain belt. However, subsequent work by \citet{Boucot2013a} and \citet{Cao2018a} has shown that evaporite deposits did form at or near the equator at times in the Phanerozoic. While some of this variability could be attributed to waxing/waning of the width of the tropical rain belt as a whole, it is important to note that there can be large deviations in local precipitation from the zonal mean due to factors such as monsoon-related precipitation, or continentality (i.e. how dispersed or amalgamated the continents are) which can lead to aridity in continental interiors. For instance, much of the Early Cretaceous low-latitude evaporite deposits were formed in basins which were located deep within arid continental interiors at the time of deposition \citep{Boucot2013a, Cao2018a}. Furthermore, in contrast to \citet{Evans2006a}, the compilation of evaporite deposits of \citet{Boucot2013a} contains sedimentary sequences in which the occurrence of evaporitic minerals is limited (e.g. to nodular evaporitic pseudomorphs). Such limited evaporitic mineral precipitation could be attributed to local processes such as periodic restriction of the basin that lead to saturation states that otherwise would not be expected for that latitude. Nevertheless, a limitation of the LIP analysis described in this study is that it does not account for deviations in local precipitation from the zonal mean (due to the infeasibility of running a global climate model at each time-step in the analysis). However, evaporite deposits are distributed bimodally about the equator in the subtropics for the vast majority of the past $\sim$420~Myr \citep{Cao2018a} and overall stability of the large-scale atmospheric circulation is predicted by climate dynamics \citep{Donohoe2017a}. Therefore, the assumption of enhanced precipitation in the tropics throughout the Phanerozoic appears to be warranted to first-order. To test the sensitivity of our results to the width of the tropical rain belt, we performed the tropical LIP area calculations with tropical rain belt widths of $\pm$10\textdegree and $\pm$20\textdegree, which yields similar results to a tropical rain belt width of $\pm$15\textdegree (Table \ref{tab:stats}).

To evaluate the relationship between Earth's climate state and tropical LIP area, we compared these areas to a compilation of the latitudinal extent of continental ice sheets through time over the Phanerozoic (\citealp{Macdonald2019a}; Fig. \ref{fig:LIP_Areas}E). The goal in doing so is to evaluate the hypothesis that there is a correlation between LIP area in the tropics and Earth's long-term climate state. The land ice record is an imperfect tracker of climate, as it is insensitive to changes in temperature during non-glacial intervals, is influenced by additional factors such as the physical geography of the continents during glacial intervals, and is potentially vulnerable to removal from the observable geologic record via erosion and burial. Nevertheless, it forms a physical record of Earth's climate through time and delineates glacial and non-glacial climate states. We take two approaches for comparison between the LIP area reconstruction and the record of ice extent. The first is to analyze the correlation between LIP area and the extent of ice away from the pole (Fig. \ref{fig:LIP_correlation}). The second is to consider the overlap between intervals of high LIP area (defined as LIP area \textgreater30\% of the maximum in a given post-emplacement model) and intervals of glacial climate (defined as ice extent \textgreater10\textdegree\xspace from the poles). This overlap approach places less emphasis on the specific magnitudes of the compiled ice extent and LIP area records.

@@

To assess the statistical significance of the correlation (or lack thereof) implied by the Pearson correlation coefficients, we applied the approach of \citet{Macdonald2019a} and simulated the four glacial episodes (Fig. \ref{fig:LIP_Areas}E) occurring at random times through the past 520~Myr, and recomputed the correlation coefficient and \% overlap between the LIP area in the tropics and the randomly timed glacial intervals for each of these 100,000 simulations (Fig. \ref{fig:LIP_correlation}). This approach accounts for the reality that spurious correlation can arise between auto-correlated data sets such as these, where each value is not independent, but is instead dependent on the previous state of the system. For the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario, 28\% of the randomly timed glacial interval simulations correlate better with the LIP area in the tropics. With an associated p-value of 0.28, the null hypothesis that glacial intervals do not correlate to LIP area in the tropics cannot be rejected. Taking this approach, none of the positive or negative correlations that emerge between the LIP area scenarios and the ice extent record are statistically significant.

@@

In the original models that proposed the `Fire and Ice' hypothesis as an explanation for the onset of the Sturtian `Snowball Earth' glaciation, chemical weathering was modeled as a function of temperature and run-off only \citep{Donnadieu2004a}. However, such an approach neglects the effects of soil shielding and regolith development in low-relief regions. Recent progress on understanding the relationships between landscapes, topography, and chemical weathering reveals that these effects are quite important \citep{Gabet2009a, Maher2014a}. As a result, more recent modeling of chemical weathering that has incorporated such processes highlights the importance of high-relief regions relative to low-relief ones for setting global weatherability \citep{West2012a, Godderis2017b}. Without active uplift, soil shielding from regolith development on low-relief LIPs could significantly decrease the local weatherability of a LIP and mute its impact on global weatherability. Furthermore, significant soil shielding can lead to a supply-limited weathering regime in which the weathering rate of the underlying bedrock becomes insensitive to kinetic and equilibrium factors such as temperature and runoff (factors that would, in the absence of soil shielding, lead to relatively high weathering rates in the tropical rain belt). In contrast, processes that lead to continued exhumation of mafic lithologies and the creation of steep topography that minimizes soil shielding, particularly in tropical regions, may exert a strong control on global weatherability and long-term climate. This interpretation underlies the hypothesis that arc-continent collisions in the tropics during the Ordovician \citep{Swanson-Hysell2017a} and the Cenozoic \citep{Jagoutz2016a} played a significant role in transitions into glacial climate states at those times - a correlation that appears robust throughout the Phanerozoic \citep{Macdonald2019a}.

A complication with the interpretation of soil shielding and limited weathering of LIPs is the rapid area decay rate inferred from the comparison of current LIP extent to estimated original extent (Fig. \ref{fig:LIP_preservation}). A couple considerations are relevant with respect to this analysis: 1) the current extent of LIP exposure is reduced in part by volcanics being covered by unconsolidated sediments (i.e. regolith development itself) in a number of the provinces; 2) the current extent of LIP exposure may be incomplete and an underestimate for some of the provinces; 3) the initial LIP extents are typically poorly constrained and are likely overestimates which could be resulting in inflated interpreted decay rates; and 4) the relationship between LIP area and volume is poorly constrained. Future efforts that improve the LIP database, such as developing better-constrained estimates of original LIP extent, constraining burial and uplift histories, and refining the timing of eruptions associated with LIPs, will improve analyses that consider the LIP record in its entirety, such as that in this contribution.

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$520~Myr. The approach of seeking to evaluate correlation between LIP area and glaciation is further complicated for Neoproterozoic Snowball Earth events because ice-albedo runaway leads to persistent global glaciation on timescales of tens of millions of years absent of continued forcing through normal carbon cycle processes until sufficient CO$_2$ to drive deglaciation builds up in the absence of silicate weathering \citep{Hoffman2017a}. Moreover, cooling past the critical threshold for rapid global glaciation may have occurred on a sub-million year timescale (e.g. \citealp{Macdonald2017a}). Nevertheless, the hypothesis of tropical LIP area associated with the ca. 720~Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth is a major motivating driver behind conducting this analysis.

@@

Together, these results indicate that the tropical area of the Franklin LIP, when compared to that of Phanerozoic as well as other Precambrian LIPs, did not have a singularly large area in the tropics. Given that similar peaks in tropical LIP area are not associated with the onset of glacial periods, additional processes beyond an increase in weatherability due to LIP area in the tropics must have been at play in the initiation of the Sturtian Snowball Earth. One such process could have been an increase in global weatherability associated with the collision and accretion of arc terranes within the present-day Arabian-Nubian Shield \citep{Park2018a}. A low-latitude continent configuration and abundant arc-continent collisions in the tropics may have led to a cool background climate, and the emplacement of the Franklin LIP may have further increased global weatherability to the point where the ice-albedo runaway could take effect. However, tropical LIP area associated with the Franklin was not uniquely high, and therefore an associated increase in global weatherability was not the sole driver of Snowball Earth onset, consistent with the results of the Phanerozoic analysis.

The temporal overlap between Franklin LIP eruptions and the initiation of Sturtian glaciation remains compelling \citep{Macdonald2010a, MacLennan2018a}. This overlap continues to suggest that other aspects of LIP emplacement, such as the injection of sulfur aerosols in the stratosphere \citep{Macdonald2017a}, could have played a role in the initiation of low-latitude glaciation. The latitude and background climate of such aerosol injections is important for their temporary effect on albedo which is maximized at low latitude, and their presence at high concentrations is pre-conditioned on eruption through sedimentary basins hosting evaporite deposits \citep{Macdonald2017a}. Sulfur aerosols associated with low-latitude LIPs may have also caused short-term cooling due to changes in albedo during the Phanerozoic, as has been suggested for the CAMP \citep{Schoene2010a}. However, because these cooling events did not result in a Snowball Earth, the climate would have returned to its background climate state within thousands of years \citep{Macdonald2017a}, which would not be observed in our analysis.

>>>>

One of the hypothesized effects of the emplacement of large igneous provinces (LIPs) is planetary cooling on million-year timescales associated with enhanced silicate weathering of the highly weatherable basalt. This study combines reconstructions of the original surface extent and emplacement ages of LIPs through the Phanerozoic, a paleogeographic model, and a paramaterization of LIP erosion to estimate LIP area in all latitudinal bands through the Phanerozoic. This analysis reveals no significant correlation between total LIP area, nor LIP area in the tropics, and the extent of continental ice sheets. The largest peaks in tropical LIP area are at times of non-glacial climate. These results suggest that changes in planetary weatherability associated with LIPs are not the fundamental control on whether Earth is in a glacial or non-glacial climate, although they could provide a secondary modulating effect in conjunction with other processes.

@@

Global weatherability is the sum of factors aside from climate itself that contribute to overall global weathering and associated CO$_2$ consumption, such as the latitudinal distribution of continents and mountain belts. On a planet with high weatherability, the CO$_2$ input from volcanism can be removed via silicate weathering at a lower atmospheric CO$_2$ concentration than on a less weatherable planet. Basaltic regions consume more CO$_2$ than regions where bedrock composition is closer to bulk continental crust because of the relatively high concentrations of Ca and Mg (that ultimately sequester carbon through precipitation as carbonate) and the high weathering rates of mafic lithologies \citep{Dessert2003a}. Furthermore, data from basaltic watersheds show that chemical weathering rates are greatest in regions with high runoff. As a result, CO$_2$ consumption in basaltic regions is most pronounced in the tropical rain belt \citep{Dessert2003a, Hartmann2009a, Hartmann2014a}.

One aspect of large igneous province (LIP) emplacement that has been hypothesized to relate to long-term global climate is the effect such emplacement of mafic lithologies could have on increasing global weatherability and driving cooling. In particular, the emplacement of LIPs in the tropics has been hypothesized to be associated with specific episodes of climatic cooling on Earth. In the Neoproterozoic Era, the emplacement of the ca. 720~Ma Franklin LIP in the tropics, in concert with elevated runoff rates associated with continental break-up, has been implicated as a major contributor to the cooling that initiated the Sturtian `Snowball Earth' \citep{Donnadieu2004b, Macdonald2010a, Cox2016a}. In the Cenozoic, the movement of the Deccan LIP into the tropical rain belt, in concert with the low-latitude emplacement of the Ethiopian Traps, has been implicated in drawing down CO$_2$ levels in the lead-up to Oligocene glaciation of Antarctica \citep{Kent2008a, Kent2013a}. More recently, \citet{Johansson2018a} used a paleogeographic model to suggest that tropical LIP area correlates with Phanerozoic climate change through comparisons to \textit{p}CO$_{2}$ proxy records.

@@

This study combines reconstructions of the original surface extent and emplacement ages of LIPs through the Phanerozoic, a paleogeographic model, and a parameterization of LIP erosion to estimate LIP area in all latitudinal bands through the Phanerozoic. We then compare these LIP area time series to the latitudinal extent of continental ice sheets - a proxy for Earth's long term climate state.

Outlines of the original surface extent of continental LIPs through the Phanerozoic (Fig. \ref{fig:LIP_map}) were slightly modified (to ensure that all currently exposed volcanic lithologies are encapsulated by the initial LIP area polygons) from the compilation of \citet{Ernst2017a} and \citet{Ernst2019a}, and emplacement ages were taken from the literature (Table \ref{tab:LIPs}). The LIP original surface extent compilation seeks to reconstruct the original surface extent of LIPs with the caveat that there can be significant uncertainty with doing so, particularly for older more deeply eroded LIPs. These polygons encapsulate all of the rocks associated with a given LIP, including dikes, sills, and layered intrusions, in addition to subaerial volcanics (Fig. \ref{fig:LIP_map}). For some LIPs, this approach may lead to an overestimate of original surface extent, given that subsurface intrusions could extend over a broader area than the surface volcanics. The polygons also assume complete surface coverage between wide-spread remnants, creating further potential for these outlines being over-estimates. However, despite these uncertainties, this approach likely provides the best estimates available of original extent for ancient LIPs. The extents of presently-exposed volcanics associated with LIPs that were used for present-day area estimates, and the sources that went into the construction of the original extent polygons by \citet{Ernst2017a} and \citet{Ernst2019a}, were taken from a number of resources including the PLATES compilation \citep{Coffin2006a} and more recent compilation efforts associated with the LIPs Reconstruction Project (\citealp{Ernst2013a}; Table \ref{tab:LIPs}).

After LIPs are emplaced, they will progressively erode. In order to account for the associated decrease in area with time, \citet{Godderis2017a} took the approach of fitting an exponential decay function to estimates of the original surface extent and the current surface extent for 5 LIPs. They used the resulting exponential decay constants to develop a first-order parameterization to estimate changing LIP area through time. We extend this approach to 19 basaltic LIPs for which there are estimates of the original surface extent of the province and the current surface extent of rocks associated with it (Fig. \ref{fig:LIP_preservation}). While there are significant uncertainties associated with the area estimates, this compilation suggests that an exponential decay function is an appropriate first-order representation of the progressive reduction in LIP area (Fig. \ref{fig:LIP_preservation}). The best-fit exponential function results in a LIP area half-life of 29~Myr. However, since we explicitly account for LIP burial separately in our area analysis (see below), we prefer to exclude from our estimation of a representative LIP area half-life the 6 of 19 LIPs which are inferred to have been partially or completely buried. This latter approach yields a slightly higher best-fit half-life of 36~Myr (Fig. \ref{fig:LIP_preservation}). Although the exponential fit to the 13 unburied LIPs is good (i.e. it yields a low root mean square error of 0.14), if each LIP is fit individually with an exponential decay function, there is variability in the estimated half-lives such that they range from shorter half-lives of $\sim$20~Myr up to $\sim$120~Myr for the Kalkarindji and Deccan Traps (Table \ref{tab:LIPs}). In the analysis of LIP area through time, we implement decay scenarios informed by this analysis: the `t$_{1/2}$ = 36~Myr' scenario uses the best-fit half-life of 36~Myr, while the `t$_{1/2}$ = 120~Myr' scenario uses the slower decay. Given that the post-emplacement weathering and erosional history of each LIP should be dependent on the tectonic and climatic setting that each LIP experiences during and after emplacement, this approach is simplistic but provides a framework for analysis. The LIP reconstructions used in this study also include pre-Phanerozoic LIPs. However, given the imposed exponential decay since emplacement, the inclusion of these LIPs does not significantly influence the calculated LIP areas through the Phanerozoic, which is the focus of this analysis.

Of the tectonic factors that could alter exposed LIP area, the most consequential is near-immediate burial of LIP volcanics that are co-located with a rift basin by sediments. There are numerous examples in the record where there is partial or complete burial of a LIP associated with rifting and thermal subsidence (Table \ref{tab:LIPs}). For example, the Afar LIP is both associated with the Ethiopian Traps which form plateau flood basalts as well as successful rifting in the region of the Red Sea that has resulted in burial (Fig. \ref{fig:LIP_map}). To account for the rapid decrease in exposed surface area that would result from burial by sediments in a rift basin, we impose two different burial scenarios for LIPs that are co-located with rifting. The `50\% burial' scenario imposes instantaneous burial of 50$\%$ of the LIP area while the `100\% burial' imposes instantaneous burial of the entire LIP as an end-member scenario. A limitation of our treatment of LIPs that are co-located with rifting is that we `bury' all of these LIPs instantly at the time of emplacement and to the same degree, when in fact the degree/timing of burial of these LIPs may vary substantially. The LIP area analysis uses all of the available combinations of the distinct decay and burial scenarios described above.

@@

The original extent LIP polygons were assigned a plate ID corresponding to a tectonic unit on Earth using the polygons of \citet{Torsvik2016a} for the Phanerozoic. The LIP polygons and tectonic units were reconstructed from 520~Ma to the present (e.g. Fig. \ref{fig:Reconstruction_Snapshots}) utilizing the paleogeographic model of \citet{Torsvik2016a} in the spin axis reference frame (anchor plate ID of 1). This paleogeographic model was updated to include revisions to Ordovician Laurentia \citep{Swanson-Hysell2017a} and Paleozoic Asia \citep{Domeier2018a}. Reconstructions and area calculations within latitude bands utilized the pyGPlates function library and custom Python scripts. The total LIP area and the LIP area reconstructed within the tropical rain belt were calculated for the various decay and burial scenarios at a resolution of 5~Myr (Fig. \ref{fig:LIP_Areas}B and C). All the data and code necessary to reproduce the analyses and figures presented in this study can be downloaded from GitHub: \url{https://github.com/Swanson-Hysell-Group/Large_Igneous_Provinces}.

We use 15\textdegree~S to 15\textdegree~N as the definition of the tropical rain belt, as those latitudes approximately correspond with a sharp increase in zonal mean precipitation to values greater than 1.0~m/yr in modern climatalogical data \citep{Kalnay1996a}. In \citet{Evans2006a}, the reconstructed paleolatitudes of basins with thick, basin-wide evaporite deposition are shown to be consistently in the subtropics throughout the Phanerozoic (and potentially beyond), suggesting that the large-scale atmospheric circulation that gives rise to intense precipitation in the tropical rain belt is stable through time. However, subsequent work by \citet{Boucot2013a} and \citet{Cao2018a} has shown that evaporite deposits did form at or near the equator at times in the Phanerozoic. While some of this variability could be attributed to waxing/waning of the width of the tropical rain belt as a whole, it is important to note that there can be large deviations in local precipitation from the zonal mean due to factors such as monsoon-related precipitation, or continentality (i.e. how dispersed or amalgamated the continents are) which can lead to aridity in continental interiors. For instance, much of the Early Cretaceous low-latitude evaporite deposits were formed in basins which were located deep within arid continental interiors at the time of deposition \citep{Boucot2013a, Cao2018a}. Furthermore, in contrast to \citet{Evans2006a}, the compilation of evaporite deposits of \citet{Boucot2013a} contains sedimentary sequences in which the occurrence of evaporitic minerals is limited (e.g. to nodular evaporitic pseudomorphs). Such limited evaporitic mineral precipitation could be attributed to local processes such as periodic restriction of the basin that led to saturation states that otherwise would not be expected for that latitude. Nevertheless, a limitation of the LIP analysis described in this study is that it does not account for deviations in local precipitation from the zonal mean (due to the infeasibility of running a global climate model at each time-step in the analysis). However, evaporite deposits, including those in which the occurrence of evaporitic minerals is limited, are distributed bimodally about the equator in the subtropics for the vast majority of the past $\sim$420~Myr \citep{Cao2018a} and overall stability of the large-scale atmospheric circulation is predicted by climate dynamics \citep{Donohoe2017a}. Therefore, the assumption of enhanced precipitation in the tropics throughout the Phanerozoic appears to be warranted to first-order. To test the sensitivity of our results to the width of the tropical rain belt, we performed the tropical LIP area calculations with tropical rain belt widths of $\pm$10\textdegree\xspace and $\pm$20\textdegree, which yields similar results to a tropical rain belt width of $\pm$15\textdegree\xspace (Table \ref{tab:stats}).

To evaluate the relationship between Earth's climate state and total and tropical LIP area, we compared these areas to a compilation of the latitudinal extent of continental ice sheets through time over the Phanerozoic (\citealp{Macdonald2019a}; Fig. \ref{fig:LIP_Areas}E). The goal in doing so is to evaluate the hypothesis that there is a correlation between LIP area in the tropics and Earth's long-term climate state. The land ice record is an imperfect tracker of climate, as it is insensitive to changes in temperature during non-glacial intervals, is influenced by additional factors such as the physical geography of the continents during glacial intervals, and is potentially vulnerable to removal from the observable geologic record via erosion and burial. Nevertheless, it forms a physical record of Earth's climate through time and delineates glacial and non-glacial climate states. We take two approaches for comparison between the LIP area reconstruction and the record of ice extent. The first is to calculate the Pearson correlation coefficient between LIP area and the extent of ice away from the pole. The second is to consider the overlap between intervals of high LIP area (defined as LIP area \textgreater30\% of the maximum in a given post-emplacement model) and intervals of glacial climate (defined as ice extent \textgreater10\textdegree\xspace from the poles). This overlap approach places less emphasis on the specific magnitudes of the compiled ice extent and LIP area records.

@@

To assess the statistical significance of the correlation implied by the Pearson correlation coefficients (or lack thereof), we applied the approach of \citet{Macdonald2019a} and simulated the four glacial episodes (Fig. \ref{fig:LIP_Areas}E) occurring at random times through the past 520~Myr, and recomputed the correlation coefficient and \% overlap between the LIP area in the tropics and the randomly timed glacial intervals for each of these 100,000 simulations (Fig. \ref{fig:LIP_correlation}). This approach accounts for the reality that spurious correlation can arise between auto-correlated data sets such as these, where each value is not independent, but is instead dependent on the previous state of the system. For the `t$_{1/2}$ = 120~Myr + 100\% burial' scenario, 72\% of the randomly timed glacial interval simulations correlate better with LIP area in the tropics. With an associated p-value of 0.72, the null hypothesis that glacial intervals do not correlate to LIP area in the tropics cannot be rejected. Taking this approach, none of the positive or negative correlations that emerge between the LIP area scenarios and the ice extent record are statistically significant (Table \ref{tab:stats}).

@@

In the original models that proposed the `Fire and Ice' hypothesis as an explanation for the onset of the Sturtian `Snowball Earth' glaciation, chemical weathering was modeled as a function of temperature and run-off only \citep{Donnadieu2004a}. However, such an approach neglects the effects of soil shielding and regolith development in low-relief regions. Recent progress on understanding the relationships between landscapes, topography, and chemical weathering reveals that these effects are important \citep{Gabet2009a, Maher2014a}. Given enough of it, soil shielding can lead to a supply-limited weathering regime in which the weathering rate of the underlying bedrock becomes insensitive to kinetic and equilibrium factors such as temperature and runoff - factors that would, in the absence of soil shielding, lead to relatively high weathering rates in the tropical rain belt. As a result, more recent modeling of chemical weathering incorporates such processes and highlights the importance of high-relief regions relative to low-relief ones for setting global weatherability \citep{West2012a, Godderis2017b}. LIPs are often emplaced in low-relief areas, and as such, without active uplift, soil shielding from regolith development on these low-relief LIPs could significantly decrease the local weatherability of a LIP and mute its impact on global weatherability. In this way, soil shielding could explain the lack of correlation between tropical LIP area and ice extent (Figs. \ref{fig:LIP_Areas} and \ref{fig:LIP_correlation}). In contrast, processes that lead to continued exhumation of mafic lithologies and the creation of steep topography that minimizes soil shielding, particularly in tropical regions, may exert a strong control on global weatherability and long-term climate. This interpretation underlies the hypothesis that arc-continent collisions in the tropics during the Ordovician \citep{Swanson-Hysell2017a} and the Cenozoic \citep{Jagoutz2016a} played a significant role in transitions into glacial climate states at those times - a correlation that appears robust throughout the Phanerozoic \citep{Macdonald2019a}.

A complication with the interpretation of soil shielding and limited weathering of LIPs is the rapid area decay rate (t$_{1/2}$ = 36~Myr) inferred from the comparison of current LIP extent to estimated original extent (Fig. \ref{fig:LIP_preservation}). A couple considerations are relevant with respect to this analysis: 1) the current extent of LIP exposure is reduced in part by volcanics being covered by unconsolidated sediments (i.e. regolith development itself) in a number of the provinces; 2) the current extent of LIP exposure may be incomplete and an underestimate for some of the provinces; 3) the initial LIP extents are typically poorly constrained and are likely overestimates which could be resulting in inflated interpreted decay rates; and 4) the relationship between LIP area and volume is poorly constrained. Future efforts that improve the LIP database, such as developing better-constrained estimates of original LIP extent, constraining burial and uplift histories, and refining the timing of eruptions associated with LIPs, will improve analyses that consider the LIP record in its entirety, such as that in this contribution.

We have focused this analysis on the Phanerozoic record given that well-constrained paleogeographic models are available for the past $\sim$520~Myr. The approach of seeking to evaluate correlation between LIP area and glaciation is further complicated for Neoproterozoic Snowball Earth events because ice-albedo runaway leads to persistent global glaciation on timescales of tens of millions of years without continued forcing through normal carbon cycle processes until sufficient CO$_2$ to drive deglaciation builds up in the absence of silicate weathering \citep{Hoffman2017a}. Moreover, cooling past the critical threshold for rapid global glaciation may have occurred on a sub-million year timescale (e.g. \citealp{Macdonald2017a}). Nevertheless, evaluating the hypothesis of tropical LIP area associated with the ca. 720~Ma Franklin LIP increasing global weatherability and contributing to the onset of the Sturtian Snowball Earth is a major motivating driver behind conducting this analysis.

@@

Together, these results indicate that the Franklin LIP, when compared to Phanerozoic as well as other Precambrian LIPs, did not have a uniquely large area in the tropics. Given that similar (and larger) peaks in tropical LIP area are not associated with the onset of glacial periods, additional processes beyond an increase in weatherability due to LIP area in the tropics must have been at play in the initiation of the Sturtian Snowball Earth. One such process could have been an increase in global weatherability associated with the collision and accretion of arc terranes within the present-day Arabian-Nubian Shield \citep{Park2018a}. A low-latitude continent configuration and abundant arc-continent collisions in the tropics may have led to a cool background climate, and the emplacement of the Franklin LIP may have further increased global weatherability to the point where the ice-albedo runaway could take effect. However, tropical LIP area associated with the Franklin was not uniquely high, and therefore an associated increase in global weatherability was not the sole driver of Snowball Earth onset, consistent with the results of the Phanerozoic analysis.
